---
title: "GSU Waiver Report"
author: "J. Crawford & S. Ridley"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document:
    df_print: paged
    theme: flatly
    highlight: haddock
    toc: yes
    toc_float: yes
---

```{r include = FALSE}

setwd("~/Graduate School Waivers")

knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE, 
                      fig.width = 10)     # Global knitr options

library(tidyr)
library(dplyr)
library(readxl)
library(scales)
library(panthr)
library(ggplot2)
library(lubridate)
library(treemapify)
library(kableExtra)                       # Load packages

options(scipen = 999)                     # Disable sci. notation

waiver <- read_excel(path = "Summer 2020 Waivers by College as of 7.15.2020.xlsx", 
                     sheet = "DB Comptroller Waiver by Term_2") %>%
  field_case(case = "snake") %>%
  rename(date = systemdate,
         pidm = sfrstcr_pidm,
         level = student_level_code,
         college = `college_description...11`,
         in_state = instatewaiver,
         out_state = outofstatewaiver) %>%
  select(date, pidm, level, college, 
         prog_code, in_state, out_state) %>%
  mutate(in_state = ifelse(is.na(in_state), 0, in_state),
         out_state = ifelse(is.na(out_state), 0, out_state))

```

<br>

# Overview

```{r}

# DETERMINE DATE POINT OR RANGE

min_date <- min(waiver$date)
max_date <- max(waiver$date)

range_dates <- ifelse(test = min_date == max_date,
                      yes = paste0("These data are current as of ", 
                                   day(min_date), " ",
                                   month(min_date, 
                                         label = TRUE, 
                                         abbr = FALSE), ", ", 
                                   year(min_date), "."),
                      no = paste0("These data range from ", 
                                  day(max_date), " ",
                                  month(min_date, 
                                        label = TRUE, 
                                        abbr = FALSE), ", ", 
                                  year(min_date), " to ",
                                  day(max_date), " ",
                                  month(max_date, 
                                        label = TRUE, 
                                        abbr = FALSE), ", ", 
                                  year(max_date), "."))

```

The following prototype is a dynamic representation of in- and out-of-state 
student waiver data at **[Georgia State University](https://www.gsu.edu/)**. The
tables and visualizations within will update with the introduction of new data.

**`r range_dates`**

<br>
<br>

# All Waivers

```{r}

# TOTALS

all_in_state <- sum(waiver$in_state, na.rm = TRUE)
all_out_state  <- sum(waiver$out_state, na.rm = TRUE)

all_in_out <- sum(c(all_in_state, all_out_state))

prop_in_state <- all_in_state / all_in_out
prop_out_state <- all_out_state / all_in_out

```

The following provides waiver totals across colleges and levels.

* Out-of-state waivers total **`r dollar(all_out_state)`** 
(**`r percent(prop_out_state, 0.01)`**)

* In-state waivers total **`r dollar(all_in_state)`** 
(**`r percent(prop_in_state, 0.01)`**)

* Total waivers equal **`r dollar(all_in_out)`**

<br>
<br>

```{r fig.height = 3}

wavers_by_state <- data.frame(Location = c("Out-of-State", "In-State"), 
                              Value = c(all_out_state, all_in_state))

ggplot(wavers_by_state, 
       aes(x = Location,
           y = Value)) + 
  geom_bar(stat = "identity",
           fill = "#0071CE", 
           width = 0.75,
           alpha = 0.75) +
  labs(title = "Total Waiver Value by Location",
       subtitle = "All Colleges, All Levels",
       caption = "Source: GSU Warehouse Data",
       x = NULL,
       y = "Value (USD)") +
  scale_y_continuous(labels = dollar) +
  coord_flip() +
  theme_minimal()

```

<br>
<br>

```{r fig.height = 2}

ggplot(wavers_by_state, 
       aes(label = Location,
           fill = Value,
           area = Value)) + 
  labs(title = "Proportion of Waive Value by Location",
       subtitle = "All Colleges, Levels, & Programs",
       caption = "Source: GSU Warehouse Data") +
  geom_treemap(show.legend = FALSE) +
  geom_treemap_text(color = "white",
                    show.legend = FALSE, 
                    padding.x = unit(0.05, 
                                     units = "inch"), 
                    padding.y = unit(0.05, 
                                     units = "inch"))

```

<br>
<br>

# Waivers by College

The following depicts total waivers aggregated by college at all levels.

<br>

```{r}

waivers_by_college <- waiver %>%
  group_by(college) %>%
  summarize(Value = sum(c(in_state, out_state))) %>%
  ungroup() %>%
  arrange(-Value) %>%
  mutate(Total = sum(Value),
         Proportion = Value / Total) %>%
  rename(College = college) %>%
  select(-Total)

waivers_by_college %>%
  mutate(Value = dollar(Value),
         Proportion = percent(Proportion, 0.01)) %>%
  kable %>%
  kable_styling(bootstrap_options = c("striped",
                                      "hover"))
  

```

<br>

```{r fig.height = 6}

ggplot(waivers_by_college, 
       aes(x = reorder(College, Value),
           y = Value)) + 
  geom_bar(stat = "identity",
           fill = "#0071CE", 
           alpha = 0.75) +
  labs(title = "Total Waiver Value by College",
       subtitle = "In-State, Out-of-State, & All Levels",
       caption = "Source: GSU Warehouse Data",
       x = NULL,
       y = "Value (USD)") +
  scale_y_continuous(labels = dollar) +
  coord_flip() +
  theme_minimal()

```

<br>
<br>

```{r fig.height = 4}

ggplot(waivers_by_college, 
       aes(label = College,
           fill = Value,
           area = Value)) + 
  labs(title = "Proportion of Total Waiver Value by College",
       subtitle = "In-State, Out-of-State, & All Levels",
       caption = "Source: GSU Warehouse Data") +
  geom_treemap(show.legend = FALSE) +
  geom_treemap_text(color = "white",
                    show.legend = FALSE, 
                    padding.x = unit(0.05, 
                                     units = "inch"), 
                    padding.y = unit(0.05, 
                                     units = "inch"))

```

<br>
<br>

## College & Location

The following itemizes waivers by college and location across levels.

<br>

```{r}

waivers_by_col_loc <- waiver %>%
  group_by(college) %>%
  summarize(in_state = sum(in_state),
            out_state = sum(out_state)) %>%
  ungroup() %>%
  mutate(prop_in_state = in_state / sum(in_state),
         prop_out_state = out_state / sum(out_state)) %>%
  arrange(-out_state) %>%
  select(college, in_state, prop_in_state, out_state, prop_out_state) %>%
  rename(College = college,
         "In-State" = in_state,
         "Out-of-State" = out_state,
         "In-State (%)" = prop_in_state,
         "Out-of-State (%)" = prop_out_state)

waivers_by_col_loc %>%
  mutate(`In-State` = dollar(`In-State`),
         `In-State (%)` = percent(`In-State (%)`, 0.01),
         `Out-of-State` = dollar(`Out-of-State`),
         `Out-of-State (%)` = percent(`Out-of-State (%)`, 0.01)) %>%
  kable %>%
  kable_styling(bootstrap_options = c("striped",
                                      "hover"))
  

```

<br>

```{r fig.height = 7}

waivers_by_col_loc_viz <- waivers_by_col_loc %>%
  select(-`In-State (%)`, 
         -`Out-of-State (%)`) %>%
  gather(key = Location, 
         value = Value, 
         -College)

waivers_by_col_loc_viz %>%
  ggplot(aes(x = reorder(College, Value),
             y = Value,
             fill = Location)) +
  coord_flip() +
  geom_bar(stat = "identity", 
           position = "dodge",
           alpha = 0.75) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Waiver Value by College & Location",
       subtitle = "All Levels",
       caption = "Source: GSU Warehouse Data",
       x = NULL,
       y = "Value (USD)") +
  theme_minimal() +
  scale_fill_manual(values = c("#00AEEF", "#0039A6"))

```

<br>
<br>

```{r}

ggplot(waivers_by_col_loc_viz, 
       aes(label = College,
           fill = Value,
           area = Value,
           subgroup = Location)) + 
  labs(title = "Proportion of Waiver Value by College & Location",
       subtitle = "All Levels",
       caption = "Source: GSU Warehouse Data") +
  geom_treemap(show.legend = FALSE) +
  geom_treemap_text(color = "white",
                    show.legend = FALSE, 
                    padding.x = unit(0.05, 
                                     units = "inch"), 
                    padding.y = unit(0.05, 
                                     units = "inch")) +
    geom_treemap_subgroup_border(color = "white", 
                               show.legend = FALSE) +
  geom_treemap_subgroup_text(color = "white", 
                             fontface = "bold",
                             show.legend = FALSE, 
                             padding.x = unit(0.1, 
                                              units = "inch"), 
                             padding.y = unit(0.1, 
                                              units = "inch"))

```

<br>
<br>

## College, Location, & Level

Waivers by college, location, and student level.

<br>

```{r}

waiver %>%
  select(college, level, in_state, out_state) %>%
  group_by(college, level) %>%
  summarize(n = n(),
            in_state = sum(in_state),
            out_state = sum(out_state)) %>%
  ungroup() %>%
  arrange(college, level, -n) %>%
  mutate(prop_in_state = in_state / sum(in_state),
         prop_out_state = out_state / sum(out_state)) %>%
  mutate(n = comma(n, 1),
         in_state = dollar(in_state),
         out_state = dollar(out_state),
         prop_in_state = percent(prop_in_state, 1),
         prop_out_state = percent(prop_out_state, 1)) %>%
  select(college:in_state, prop_in_state, 
         out_state, prop_out_state) %>%
  rename(College = college,
         Level = level,
         Count = n,
         `In-State` = in_state,
         `Out-of-State` = out_state,
         `In (%)` = prop_in_state,
         `Out (%)` = prop_out_state) %>%
  kable %>%
  kable_styling(bootstrap_options = c("striped",
                                      "hover"))
  
  

```

<br>

```{r fig.height = 8}

waiver %>%
  select(college, level, in_state, out_state) %>%
  rename(`In-State` = in_state,
         `Out-of-State` = out_state) %>%
  gather(key = location, 
         value = value, 
         -college, 
         -level) %>%
  group_by(college, level, location) %>%
  summarize(value = sum(value)) %>%
  ungroup() %>%
  select(college, location, level, value) %>%
  arrange(college, level, location, value) %>%
  rename(College = college,
         Location = location, 
         Level = level,
         Value = value) %>%
  ggplot(aes(x = reorder(College, Value),
             y = Value,
             fill = Location)) +
  coord_flip() +
  geom_bar(stat = "identity", 
           position = "dodge",
           alpha = 0.75) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Waiver Value by College & Location",
       subtitle = "Faceted by Level",
       caption = "Source: GSU Warehouse Data",
       x = NULL,
       y = "Value (USD)") +
  theme_minimal() +
  facet_wrap(~Level) +
  theme(legend.position = "bottom") +
  scale_fill_manual(values = c("#00AEEF", "#0039A6"))

```

<br>
<br>

# Waivers by College & Program

Proportion of total tuition value by programs in each college.

<br>
<br>

## All Colleges & Programs

Proportions of total tuition value across all colleges and programs.

<br>

```{r out.width = "100%", fig.height = 10}

waiver %>%
  select(college, prog_code, in_state, out_state) %>%
  group_by(college, prog_code) %>%
  summarize(in_state = sum(in_state),
            out_state = sum(out_state),
            value = in_state + out_state) %>%
  ungroup() %>%
  select(-in_state, -out_state) %>%
  ggplot(aes(area = value,
             fill = value,
             label = prog_code, 
             subgroup = college)) +
  geom_treemap(show.legend = FALSE) +
  geom_treemap_subgroup_border(color = "white", 
                               show.legend = FALSE) +
  geom_treemap_subgroup_text(color = "white", 
                             fontface = "bold",
                             show.legend = FALSE, 
                             padding.x = unit(0.1, 
                                              units = "inch"), 
                             padding.y = unit(0.1, 
                                              units = "inch")) +
  geom_treemap_text(color = "white",
                    show.legend = FALSE, 
                    padding.x = unit(0.05, 
                                     units = "inch"), 
                    padding.y = unit(0.05, 
                                     units = "inch")) +
  labs(title = "Proportion of Total Waiver Value by College & Program",
       subtitle = "All Programs, Locations, & Levels",
       caption = "Source: GSU Warehouse Data")

```

<br>
<br>

## College of Arts & Sciences

```{r}

coas_sum <- waiver %>%
  select(college, prog_code, in_state, out_state) %>%
  group_by(college, prog_code) %>%
  summarize(n = n(),
            in_state = sum(in_state),
            out_state = sum(out_state),
            value = in_state + out_state) %>%
  ungroup() %>%
  select(-college, -in_state, -out_state) %>%
  summarize(mean_n = mean(n),
            median_n = median(n),
            max_n = max(n),
            mean_value = mean(value),
            median_value = median(value),
            max_value = max(value))

```

Total waiver value by program for the College of Arts & Sciences.  

<br>

For unique students in each program:

* Average of **`r round(coas_sum$mean_n, 2)`**
* Median of **`r coas_sum$median_n`**
* Maximum of **`r coas_sum$max_n`**
 
<br>

For total waiver value in each program:

* Average of **`r dollar(coas_sum$mean_value)`**
* Median of **`r dollar(coas_sum$median_value)`**
* Maximum of **`r dollar(coas_sum$max_value)`**

<br>

#### Top 15 Total Waiver Values by Program, COAS

```{r}

waiver_coas <- waiver %>%
  select(college, prog_code, in_state, out_state) %>%
  group_by(college, prog_code) %>%
  summarize(n = n(),
            in_state = sum(in_state),
            out_state = sum(out_state),
            value = in_state + out_state) %>%
  ungroup() %>%
  select(-in_state, -out_state) %>%
  filter(college == "College of Arts & Sciences") %>%
  arrange(-value)

waiver_coas %>% 
  select(-college) %>%
  rename(Program = prog_code,
         Value = value,
         Count = n) %>%
  mutate(`Count (%)` = percent(Count / sum(Count), 0.1),
         `Value (%)` = percent(Value / sum(Value), 0.1),
         Value = dollar(Value)) %>%
  select(Program, Count, `Count (%)`, Value, `Value (%)`) %>%
  head(15) %>%
  kable %>%
  kable_styling(bootstrap_options = c("striped",
                                      "hover"))

```

<br>
<br>

```{r fig.height = 7, out.width = "90%"}

waiver_coas %>%
  head(15) %>%
  rename(Program = prog_code,
         Value = value) %>% 
  ggplot(aes(x = reorder(Program, Value),
           y = Value)) + 
  geom_bar(stat = "identity",
           fill = "#0071CE", 
           alpha = 0.75) +
  labs(title = "Top 15 Total Waiver Values by Program, COAS",
       subtitle = "All Programs, Locations, & Levels",
       caption = "Source: GSU Warehouse Data",
       x = NULL,
       y = "Value (USD)") +
  scale_y_continuous(labels = dollar) +
  coord_flip() +
  theme_minimal()

```

<br>
<br>

```{r out.width = "100%"}

waiver_coas %>%
  ggplot(aes(area = value,
             fill = value,
             label = prog_code, 
             subgroup = college)) +
  geom_treemap(show.legend = FALSE) +
  geom_treemap_subgroup_border(color = "white", 
                               show.legend = FALSE) +
  geom_treemap_text(color = "white",
                    show.legend = FALSE, 
                    padding.x = unit(0.05, 
                                     units = "inch"), 
                    padding.y = unit(0.05, 
                                     units = "inch")) +
  labs(title = "Proportion of Total Waiver Value, COAS",
       subtitle = "All Programs, Locations, & Levels",
       caption = "Source: GSU Warehouse Data")

```





